Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-04-02 jss/EpiCompare-paper/EpiCompare-paper.Rmd
update to 2.2 and comparisons and assessment in section 3.

##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #beyond-r0-sir}

Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted. \ben{\sout{In this section we will present} time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown \shannon{or have more than three epidemic states}. \sout{We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }

\ben{New paragraph} Viewing epidemics as filamental trajectories provide \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner. For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{are we missing a few words in this definition?}. This \shannon{representation induces}\sout{ also creates} a natural distance between this type of representation between epidemics, specifically:
\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\]
where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$. This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. This \shannon{what is this?} can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before... \shannon{I don't think we have but am wondering if we're getting in the weeds}}} (by making all values the proportion of the population in the given state).

\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}]. In these settings we go a step farther and treat epidemics more like geometric filaments \shannon{(i.e. filamental trajectories without an ordering of points)} than filamental trajectories. In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above pseudo-density estimates or depth estimates). These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf \shannon{why Haussdorf specifically?} distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region. All these geometric structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).

\ben{[Ben says: Old recommendations:
to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset. Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}

\shannon{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}

@@

\ben{[This still needs to be worked on - just a reminder.]}

@@

As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of potential for time-invariant tools to help compare and assess epidemics and models/simulations. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g. means squared error, AIC) and focus on assessing the structural information the models capture. This approaches work well on models where online one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again). Epidemics are complex objects, and we provide tools to create prediction regions out of simulated epidemics. For 3-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations (where the simulations treated as filaments). In \pkg{Epicompare} we also provide these prediction regions for epidemic models with with more than 3 states. With `create_convex_hull_structure` and ``create_delta_ball_structure` that will create different geometric representations of prediction regions and an assoicated `contained` method that assesses if a new epidemic (seen as a filament) is contained in the prediction region.

We also provide functions to calculate the extremeness of a true epidemic compared to simulated epidemics through the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions like  `distance_psuedo_density_function` can calculate a psuedo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function` and `local_distance_depth_function` provide depth scores that suggest how geometrically central an epidemic is to simulations.

>>>>

##  Time-invariant analysis beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #sec:beyond-r0-sir}

Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted. Time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown or have more than three epidemic states.

\ben{New paragraph} Viewing epidemics as filamental trajectories provide \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner. \ben{For epidemics that have ended, one way to examine their filamental trajectories is to redefine the trajectory as a finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).} \sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{[Shannon says: are we missing a few words in this definition?]}}. This representation induces a natural distance between this type of representation between epidemics, specifically:
\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\]
where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$. This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness estiamtors can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\ben{I'm not sure we've talked about this before... \shannon{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.

\ben{New paragraph:} \ben{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If one is interested in understanding epeidmeics through a single cycle of their outbreak} (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}]. In these settings we go a step farther and treat epidemics more like geometric filaments \shannon{(i.e. filamental trajectories without an ordering of points)} than filamental trajectories. In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above pseudo-density estimates or depth estimates). These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using \ben{many set difference distances \sout{the Haussdorf \shannon{why Haussdorf specifically?} distance}} as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region. All these geometric structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).

@@

@@

As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of potential for time-invariant tools to help compare and assess epidemics and models/simulations. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g. means squared error, AIC) and focus on assessing the structural information the models capture. This approaches work well on models where online one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again)\footnote{\ben{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}. Epidemics are complex objects, and we provide tools to create prediction regions out of simulated epidemics. For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations (where the simulations treated as filaments). In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure` and `create_delta_ball_structure` create different geometric representations of prediction regions for any state-based model. For both of these geometric objects, we provide functions to check if a path is contained (`contained`) and the ability to assess the Hassdorf distance between prediction regions based on simulations from different model (`hassdorf_dist`).  \ben{[Ben says: should we discuss the details of `geom_prediction_bands` more?]}

We also provide functions to calculate the extremeness of a true epidemic compared to simulated epidemics through the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions like  `distance_psuedo_density_function` can calculate a psuedo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function` and `local_distance_depth_function` provide depth scores that suggest how geometrically central an epidemic is to simulations. \ben{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}

### 2021-04-02 jss/EpiCompare-paper/EpiCompare-paper.tex
Auto stash before merge of "master" and "origin/master"
update to 2.2 and comparisons and assessment in section 3.

\subsection[Beyond R0 and SIR]{\textcolor{violet}{Time-invariant analysis}
Beyond \(R_0\) and Kermack's and McKendrick SIR
Models\footnote{Probably will need to change this title...}}\label{beyond-r0-sir}

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space allows us
to better compare the overall structure of the epidemic and see how the
population was directly impacted.
\textbackslash ben\{\sout{In this section we will present}
time-invariant tools \sout{that} can be useful even when the underlying
generative model for the epidemic is unknown
\textcolor{orange}{or have more than three epidemic states}.
\sout{We will also present how this tools can be used when the epidemics have more than 3 states.}
\sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }

\textcolor{violet}{New paragraph} Viewing epidemics as filamental
trajectories provide \sout{a lot }new ways to compare and examine
epidemics in a time-invariant manner. For epidemics that have ``played''
themselves out, one way to represent their filamental trajectories to
avoid \textcolor{orange}{confusion stemming from}\sout{impacts of}
temporal structure is to define them as a sequence of points their
trajectory with equi-distance between each point
\textcolor{orange}{are we missing a few words in this definition?}. This
\textcolor{orange}{representation induces}\sout{ also creates} a natural
distance between this type of representation between epidemics,
specifically: \[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
fraction of \(|\psi_1|\) distance away from the start of \(\psi_i\).
This distance is naturally time-invariant, and can be plugged into
multiple distance-based assessment tools to examine the overall
extremeness of points, including psuedo-density estimators and
depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. This
\textcolor{orange}{what is this?} can be very useful when comparing
between a set of simulation epidemics and the true epidemic, and does
not constrain the number of states of the model, though we recommend
projecting the points into the unit
simplex\footnote{\textcolor{violet}{I'm not sure we've talked about this before... \textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}
(by making all values the proportion of the population in the given
state).

\textcolor{violet}{New paragraph:} If the set of epidemics that one is
examining have only gone through a single cycle of the outbreak (before
the population of individuals have become susceptible again), then
additional time-invariant tools, including prediction regions can we
leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done?}\}}.
In these settings we go a step farther and treat epidemics more like
geometric filaments
\textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
than filamental trajectories. In \pkg{EpiCompare} we create prediction
regions that contain a the top (\(1-\alpha\)) proportion of simulated
curves by defining geometric regions defined by the union of small
filaments around the subset of simulations (subset by measures like the
above pseudo-density estimates or depth estimates). These regions look
at where in the state-space we expect the epidemic to traverse, and we
can compare prediction regions defined by different models using the
Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance as
well as examining how well the truth epidemic matches the simulations by
examining if the epidemic's trajectory lies within the prediction
region. All these geometric structures and distances notations apply to
epidemics with any number of states, and at the end of Section
\ref{sec:overview} we \sout{also} highlight how these prediction regions
can aid in visual comparisons for epidemics with 3 states (like the SIR
models).

\textcolor{violet}{[Ben says: Old recommendations:
to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset. Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}

\textcolor{orange}{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}

@@

\textcolor{violet}{[This still needs to be worked on - just a reminder.]}

As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
potential for time-invariant tools to help compare and assess epidemics
and models/simulations. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for models that extend beyond the
standard performance metrics (e.g.~means squared error, AIC) and focus
on assessing the structural information the models capture. This
approaches work well on models where online one ``cycle'' of the
epidemic has occurred (no recovered individuals have been susceptible
again). Epidemics are complex objects, and we provide tools to create
prediction regions out of simulated epidemics. For 3-state epidemic
models, we provide the \texttt{ggplot}/\texttt{ggtern} extension
\texttt{geom\_prediction\_band()} which creates a prediction region
around the top \(1-\alpha\) proportion of the simulations (where the
simulations treated as filaments). In \pkg{Epicompare} we also provide
these prediction regions for epidemic models with with more than 3
states. With \texttt{create\_convex\_hull\_structure} and
`\texttt{create\_delta\_ball\_structure} that will create different
geometric representations of prediction regions and an assoicated
\texttt{contained} method that assesses if a new epidemic (seen as a
filament) is contained in the prediction region.

We also provide functions to calculate the extremeness of a true
epidemic compared to simulated epidemics through the equi-distance
filamental trajectory representation as mentioned in Section
\ref{sec:beyond-r0-sir}. Specifically, functions like
\texttt{distance\_psuedo\_density\_function} can calculate a
psuedo-density estimate of the true epidemic relative to simulated ones.
Functions \texttt{distance\_depth\_function} and
\texttt{local\_distance\_depth\_function} provide depth scores that
suggest how geometrically central an epidemic is to simulations.

>>>>

\subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
and Kermack's and McKendrick SIR
Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space allows us
to better compare the overall structure of the epidemic and see how the
population was directly impacted. Time-invariant tools \sout{that} can
be useful even when the underlying generative model for the epidemic is
unknown or have more than three epidemic states.

\textcolor{violet}{New paragraph} Viewing epidemics as filamental
trajectories provide \sout{a lot }new ways to compare and examine
epidemics in a time-invariant manner.
\textcolor{violet}{For epidemics that have ended, one way to examine their filamental trajectories is to redefine the trajectory as a finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).}
\sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
This representation induces a natural distance between this type of
representation between epidemics, specifically: \[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
fraction of \(|\psi_1|\) distance away from the start of \(\psi_i\).
This distance is naturally time-invariant, and can be plugged into
multiple distance-based assessment tools to examine the overall
extremeness of points, including psuedo-density estimators and
depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estiamtors can be very useful when comparing between a set of simulation
epidemics and the true epidemic, and does not constrain the number of
states of the
model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before... \textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.

\textcolor{violet}{New paragraph:}
\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If one is interested in understanding epeidmeics through a single cycle of their outbreak}
(before the population of individuals have become susceptible again),
then additional time-invariant tools, including prediction regions can
we leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done?}\}}.
In these settings we go a step farther and treat epidemics more like
geometric filaments
\textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
than filamental trajectories. In \pkg{EpiCompare} we create prediction
regions that contain a the top (\(1-\alpha\)) proportion of simulated
curves by defining geometric regions defined by the union of small
filaments around the subset of simulations (subset by measures like the
above pseudo-density estimates or depth estimates). These regions look
at where in the state-space we expect the epidemic to traverse, and we
can compare prediction regions defined by different models using
\textcolor{violet}{many set difference distances \sout{the Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance}}
as well as examining how well the truth epidemic matches the simulations
by examining if the epidemic's trajectory lies within the prediction
region. All these geometric structures and distances notations apply to
epidemics with any number of states, and at the end of Section
\ref{sec:overview} we \sout{also} highlight how these prediction regions
can aid in visual comparisons for epidemics with 3 states (like the SIR
models).

@@

As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
potential for time-invariant tools to help compare and assess epidemics
and models/simulations. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for models that extend beyond the
standard performance metrics (e.g.~means squared error, AIC) and focus
on assessing the structural information the models capture. This
approaches work well on models where online one ``cycle'' of the
epidemic has occurred (no recovered individuals have been susceptible
again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}.
Epidemics are complex objects, and we provide tools to create prediction
regions out of simulated epidemics. For three-state epidemic models, we
provide the \texttt{ggplot}/\texttt{ggtern} extension
\texttt{geom\_prediction\_band()} which creates a prediction region
around the top \(1-\alpha\) proportion of the simulations (where the
simulations treated as filaments). In \pkg{EpiCompare} we also provide
these prediction regions for epidemic models with with more than three
states. The functions \texttt{create\_convex\_hull\_structure} and
\texttt{create\_delta\_ball\_structure} create different geometric
representations of prediction regions for any state-based model. For
both of these geometric objects, we provide functions to check if a path
is contained (\texttt{contained}) and the ability to assess the Hassdorf
distance between prediction regions based on simulations from different
model (\texttt{hassdorf\_dist}).

We also provide functions to calculate the extremeness of a true
epidemic compared to simulated epidemics through the equi-distance
filamental trajectory representation as mentioned in Section
\ref{sec:beyond-r0-sir}. Specifically, functions like
\texttt{distance\_psuedo\_density\_function} can calculate a
psuedo-density estimate of the true epidemic relative to simulated ones.
Functions \texttt{distance\_depth\_function} and
\texttt{local\_distance\_depth\_function} provide depth scores that
suggest how geometrically central an epidemic is to simulations.
\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}
